Prosper Loan Data Analysis by Oliver Phipps
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

setwd("C:/Users/phipp/Dropbox/7.Courses/DAND/Term Two/Final Project Submissions/Explore and Summarize Data")

library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
library(grid)
library(ggthemes)


```

>In this analysis I am going to look at the Prosper Loan Data - I have imported the relevant libraries and I will first take a look at the data.

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('prosperLoanData.csv')

#view the headings
str(df)

#view stats about the information
summary(df)

```

> This data set contains 113,937 loans with multiple variables on each loan, in this analysis we will look at a number a different variables to try and understand more about the data.


# Univariate Plots Section

> The below shows a summary of some of the variables that we will explore in our analysis and helps us get a preliminary understanding.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}


#view a histogram on a variety of data points within the data
histogram <- function(var, xlabel) {
    return(qplot(x = var, data = df, xlab = xlabel))
}

grid.arrange(histogram(df$Term, 'Term'),
histogram(df$BorrowerRate, 'Borrower Rate'),
histogram(df$CreditScoreRangeLower, 'Credit Score Range Lower'),
histogram(df$CreditScoreRangeUpper, 'Credit Score Range Upper'),
histogram(df$BankcardUtilization, 'Bankcard Utilization'),
histogram(df$DebtToIncomeRatio, 'Debt To Income Ratio'),
histogram(df$LoanOriginalAmount, 'Loan Original Amount'),
histogram(df$MonthlyLoanPayment, 'Monthly Loan Payment'),
histogram(df$LP_InterestandFees, 'Total Interest and Fees'),
ncol = 3)

```

>Now lets look at two variables side by side that I we would typically consider linked, the loan amounts and loan payments

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots2}

#view a histogram of two data points side by side
grid.arrange(qplot(x = LoanOriginalAmount, 
                   data = df, binwidth=1000),
             qplot(x = MonthlyLoanPayment, 
                   data = df, binwidth = 50), ncol = 2)
```

>We can see that these are both right skewed

>Now lets look at information on Loan amounts by state, to see if these differ, we have intentionally picked out two states that are very different in terms of wealth per capita

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots3}

#view two more histograms, this time we are filtering by state and looking at the same statistic in two different states
grid.arrange(qplot(x = LoanOriginalAmount, 
                   data = subset(df,  BorrowerState == 'CT'), 
                   ylab = 'Count for State = Connecticut', 
                   binwidth=1000),
             qplot(x = LoanOriginalAmount, 
                   data = subset(df,  BorrowerState == 'WV'), 
                   ylab = 'Count for State = West Virginia', 
                   binwidth=1000), ncol = 2)
```

>We can see above that even though the population of these states are very different in terms of wealth, they are quite similiar if we look at the amount ehey are borrowing.

>Now lets look at some catergorical variables
>First we are looking at the counts of homeowners vs non homeowners to see if they prefer to offer to loanee's with a home to use as collateral.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots4}

#pie chart to show homeowner information
ggplot(df, aes(x = IsBorrowerHomeowner)) + geom_bar() + 
    geom_text(stat='count', aes(label=..count..), vjust=1.5)
```

>We can see above that there is roughly the same amount of each

>Now lets look at the count of Loanee's by there Income Range, the reason behind this is so that we can see if there is a preference for the company when offering loans or if they specialise in offering to a certain type of Loanee.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots5}

#pie chart to show income range information
ggplot(df, aes(x = IncomeRange)) + geom_bar() + 
    geom_text(stat='count', aes(label=..count..), vjust=1.5)
```

>We can see here that it is quite distributed, going forward we will sort this data so it is ordered correctly

>Finally lets look at Prosper Rating, the lower the number the more risky the Loanee is deemed to be

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots6}

#pie chart to show Prosper Rating information
ggplot(df, aes(x = ProsperRating..numeric.)) + geom_bar() + 
    geom_text(stat='count', aes(label=..count..), vjust=1.5)
```

>This is normally distributed, with maybe a slightly left skew. This reason behind looking at this was to understand if the company only offers loans to certain types of people.

# Univariate Analysis


### What is the structure of your dataset?

> The data has 113,937 loans and 81 variables for each, including some that are shown above, and also some regional  data which we can use to investigate more regional trends.

### What is/are the main feature(s) of interest in your dataset?

> There are many aspects of this dataset and I would not say that there is one main feature. One thing that I would like to investigate however would be regarding loans and certain income bands and how the Loan amount would diffe and therefore I would take the Loan Amount to be my main point of interest.

### Did you create any new variables from existing variables in the dataset?

> I have not yet created any new variables although throughout my analysis creating new variables, or sorting factored variables, may be neccesary to get desired results.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

> This dataset does have some missing values for some of the variables, but the ones that I am interested in contain enough information that I feel confortable using this data without tidying. Although I am not familiar with US credit scores, I would have suspected some more variation in the scores however that may be due to a policy or some criteria of the lender.

> We can also see that on the Orignal Loan Amount, there are spikes around more standard multiples (e.g. 10k, 20k) and less loans that have a more specific amount. This may suggest that Loanee's are perhaps rounding their requirements up or down, or that the Loaner is more likely to offer standard amounts.

>We can see in the summary chart that most are skewed, this would be expected in this dataset as we are dealing with a number of Loanees with vastly different financial situation and needs. If we look at the state specific charts we can see that they look more similiar than may be expected for two states that are very different in terms of per capita income.


# Bivariate Plots Section



```{r echo=FALSE, Bivariate_Plots1}

#this will produce a histogram of two different data points overlayed on top of each other
ggplot(df, aes(x=LoanOriginalAmount, fill=factor(IsBorrowerHomeowner))) + 
  geom_density(alpha=0.2)

```

> The above gives us some insight as to the size of loans people take out if they own homes, and if they do not. We can see that non-homeowners take a out more smaller loans whilst people who own a home take out more. The inflection point here is around 10,000. The reason for us to look at this was to see if Home Owners may borrow more for maintenence to their property or if they borrow less as they could be perceived to be more wealthy.


```{r echo=FALSE, Bivariate_Plots2}

#this will filter re-arrange the order of how you can view the data point which is catergorical variable
df$IncomeRange = factor(df$IncomeRange,levels(df$IncomeRange)[c(1,2,4,5,6,3)])

#this will plot another overlay graph but omit anything in which the income range is not displayed
ggplot(subset(df,  IncomeRange != 'Not displayed'), 
       aes(x=LoanOriginalAmount, fill=factor(IncomeRange))) + 
  geom_density(alpha=0.2)


```

> We know that there are typically restrictions around what you can borrow depending on what you earn, but it is interesting to see this played out in the above plot. We can again see these spikes played out around round multiples to borrow (e.g. 15k shows a number of Income Ranges that are peaking.) The most interesting thing here it the prevalence of a $0 range. This may either be due to a data inputting error or around the use of collateral for loans. In this data we were using factor variables and so I used R to put these into order and also removed 'not displayed' and 'not employed'. Below are some statistics drawn from the data used in the graph above. I think this is interesting and the reason behind this was too see if there was a dramatic difference in what different people with differnt income ranges may borrow.



```{r echo=FALSE, Bivariate_Plots3}

#here we are viewing some stats on the income range data point and loan amounts
by(df$LoanOriginalAmount, factor(df$IncomeRange), summary)

```

```{r echo=FALSE, Bivariate_Plots4}
#create a new dataframe that removes top 1% so that the graph plots correctly
df1 <- df[df$StatedMonthlyIncome < quantile(df$StatedMonthlyIncome, 0.99),]

#this will plot a graph that looks at only Conneticut and looks at income compared to loan payments
#we use stat_smooth to look get a trend line
ggplot(subset(df1,  BorrowerState == 'CT'), 
       aes(StatedMonthlyIncome, MonthlyLoanPayment)) + 
  geom_point(alpha=0.1) + 
  stat_smooth(method = "lm", formula = y ~ x)

```

>The graph above looks at Loan Payments and Montly Income. The underlying data excludes the top 1% of earners and this chart only looks at residents of Connecticut. The reasoning behind this is due to the amount of entries, using subsets helps give us a better idea of trends. We can see that as expected, the more that is earned a month the higher than loan payments typically are. This is interesting to see as we can understand what a lower income state earnts a month compared to what they pay for loans.



```{r echo=FALSE, Bivariate_Plots5}
#here we view the same graph but look at West Virgina
ggplot(subset(df1,  BorrowerState == 'WV'), aes(StatedMonthlyIncome, 
                                                MonthlyLoanPayment)) + 
  geom_point(alpha=0.2) + 
  stat_smooth(method = "lm", formula = y ~ x)

```

>The above chart show similiar data to the one before it, but now we are looking at residents of West Virgina. The reason behind this is that it allowed us to compare against the information for Conneticut.

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

>As has been commented above, we can see that homeowners typically borrower more, as do those who are in higher Income Ranges. It would be interesting to do more future analysis around the $0 salary as we can see that this has a higher mean borrowering amount that the next two Income Range buckets. As expected we saw that the Monthly Payments are positively correlated with Monthly Income.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
 
> When looking at the subsets of the monthly Income vs Loan repayments it was interesting to see some of the outliers where the Loan amount being paid back is way of the trend line, for example in Conneticut where the monthly income was just under 5000, but the repayments were above 1500 per month.

# Multivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots1}

#here we look at information based on income range in conneticut only, it is colored based on the income range
ggplot(subset(df1,  BorrowerState == 'CT', IncomeRange != 'NA'), 
       aes(x=StatedMonthlyIncome, y=MonthlyLoanPayment,
           color=factor(IncomeRange))) + 
    
  geom_point() + 
  scale_color_brewer()

```

>The above shows us how monthyly income in Connecticut compares to Loans, but also breaks it out by Income Range. This is interesting as it allows to easily see (via colours) the outliers in different groups, and that was the reason for this graph.

>The below chart is looking at DC and it shows us credit scores vs income ranges. The idea behind this is to see if we are see if trends in Income Ranges relate to credit scores.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots2}


#the chart uses different size circles and different colours to short information
#the chart 3 is then used with a theme package to alter how it appears


ggplot(subset(df, BorrowerState == 'DC' 
                        & CreditScoreRangeLower > 500), 
       aes(x=LoanOriginalAmount, y=CreditScoreRangeLower,
           color=factor(IncomeRange))) + 
    
    geom_point() + 
    scale_color_brewer()

```


>Below is a breakdown of statistics for Monthly Loan Payment by Income Range

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots3}

#here are some stats on loan payment data when looking at income ranges
by(df$MonthlyLoanPayment, factor(df$IncomeRange), summary)

```



>Here we can again see the Monthly Income and Payments but we have added colour to denote what income range the Loanee falls within. We can see the clear boundaries and we can more clearly in what range the anomalies fall. Again, this is only looking at the state of Conneticut.


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots4}

#here we look at data using the Prosper Rating annd instead of chaing colour we vary the size of the circle
#the size of the circle denotes different Prosep Ratings
ggplot(subset(df1,  BorrowerState == 'CT', IncomeRange != 'NA'), 
       aes(x=StatedMonthlyIncome, y=MonthlyLoanPayment,
           color=factor(ProsperRating..numeric.))) + 
    
    geom_jitter() + 
    scale_color_brewer()

```

> Again looking only at the state of Conneticut, we have added in Prosper rating which is used and affects the size of each point. Prosper rating is used to categorise each Loanee by level of risk. The lower the number (or smaller the point in the plot) the higher risk they are. We can see above an example of this clearly, where a Loanee with a prosper rating of 1 has monthly income of just under 20,000, but rapyments of lower than 500. It is possible that the Prosper rating was the cause of such small repayments relative to monthly income. The reason behind this was that it helps show us info around how the Prosper rating may affect the Income vs Loan data.  The reason that we have use 'jitter' above is that it helps add some noise to the data to help prevent overplotting, it also helps us see trends more clearly.

>Below is a breakdown of statistics for Monthly Loan Payment by Prosper Rating

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots5}

#this shows stats on Prosper Ratings and the size of monthly loan payments
by(df$MonthlyLoanPayment, factor(df$ProsperRating..numeric.), summary)

```

>Below is a breakdown of statistics for Monthly Income by Prosper Rating

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots6}

#this shows stats on Prosper Ratings and the size of monthly income
by(df$StatedMonthlyIncome, factor(df$ProsperRating..numeric.), summary)

```


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots7}

#this shows different credit scores based on Prosper rating, with colours denoting different ratings
ggplot(subset(df1,  CreditScoreRangeLower > 500), 
       aes(x=LoanOriginalAmount, y=CreditScoreRangeLower,
           color=factor(ProsperRating..numeric.))) + 
    
    geom_jitter() + 
    scale_color_brewer()

```

>This chart now includes all data from the dataset. We are looking at the TOtal Loan Amount, the Credit Score and the Prosper Rating. The trends that are most obvious to start with are that there are higher prospect ratings in the top left of the chart, with the lower amounts and higher credit scores. This obviously makes sense. This helps show us how close the Prosper rating is affected byt Credit Rating via colours. The reason that we have use 'jitter' above is that it helps add some noise to the data to help prevent overplotting, it also helps us see trends more clearly.

>The below information shows us some of what we have already seen in terms of income range and prosper rating, but this shows us all the information together in one chart. The reason behind this is so that we can see how the variables all relate to each other.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots8}


#the chart uses different size circles and different colours to short information
#the chart 2 is then used with a theme package to alter how it appears
ggplot(subset(df,  BorrowerState == 'DC'), 
       aes(x=StatedMonthlyIncome, y=MonthlyLoanPayment,
            color=factor(IncomeRange),
           size=ProsperRating..numeric.)) + 
    
    geom_point(alpha=0.35) + 
    scale_color_brewer()


```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

>Some of the relationships have been discussed above, one which we can see is that on lower incomes the monthly payments and monthly income are much more closely knit, but on the larger incomes there is more disparity. We can also see from the third chart above that there seems to be limits for Loan Amount based on the credit score of the Loanee.

------

# Final Plots and Summary


```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

#we are using 'chart 1' to create the chart, which shows the icnome range data and loan amounts
#the chart is an overlay
#the chart 1 is then used with a theme package to alter how it appears
chart1 <- ggplot(subset(df,  IncomeRange != 'Not displayed'), 
                 aes(x=LoanOriginalAmount, fill=factor(IncomeRange))) + 
  geom_density(alpha=0.2) + 
    labs(title = 'Density Plot of Loan Amount by Income Range',
         x = 'Original Loan Amount',
         y = 'Density',
         fill = 'Income Ranges')

chart1 + theme_economist(base_family="Verdana") +
  scale_colour_economist()

```

### Density Plot of Loan Amount by Income Range
> This plot shows the Loan Amount of differing Income ranges. I think that this plot shows the habits or various Loanee's well as you can understand that that with a lower income may use small loans to pay for things like bills or unexpected accidents, as they may not have much saved to cover these costs. It also shows the general borrowing habits of higher earnings, who cannot only borrow more, but may be likely to be using the resulting loan to purchase more expensive goods. The reason that we are showing the difference in borrowing habits by Income Range and we can see if these were linked.


```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

#the df2 crates a new dataframe that excludes any lines in the Income range data points that is blank
#we are using 'chart 2' to create the chart, which shows monthly income vs loan payments
#the chart uses different size circles and different colours to short information
#the chart 2 is then used with a theme package to alter how it appears
df2 <- df1[complete.cases(df1$IncomeRange), ]

chart2 <- ggplot(subset(df2,  BorrowerState == 'DC'), 
       aes(x=StatedMonthlyIncome, y=MonthlyLoanPayment,
            color=factor(IncomeRange),
           size=ProsperRating..numeric.)) + 
    
    geom_point(alpha=0.35) + 
    scale_color_brewer() + 
    labs(title = 'Monthly Income vs Loan Payment in D.C.',
         x = 'Monthly Income (top 1% Removed)',
         y = 'Monthly Loan Payment',
         color = 'Income Ranges',
         size = 'Prosper Rating')

chart2 + theme_economist(base_family="Verdana") +
  scale_colour_economist() + theme(legend.position="right")

```

### Monthly Income vs Loan Payment in Washington, D.C.
>This graph combines two that we saw earlier buy looks at Washington, D.C. instead. It gives more information about the Prosper rating and income range vs Monthly Payments and Income. We can see the same trend of lower incomes typically being bunched closer together with the higher earners having a lot more variation in terms of the Income vs Payments. It is also not neccesarily the case the more you earn the higher your rating. After looking at income data above it was helpful for us to see this information plotted again but now looking at the Income vs Loan Payment data.



```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}

#df3 creates a new dataframe that excludes any lines where the Prosper Rating is blank
#we are using 'chart 3' to create the chart, which shows the icnome range data and loan amounts
#the chart uses different size circles and different colours to short information
#the chart 3 is then used with a theme package to alter how it appears
df3 <- df2[complete.cases(df1$ProsperRating..numeric.), ]

chart3 <- ggplot(subset(df3, BorrowerState == 'DC' 
                        & CreditScoreRangeLower > 500), 
       aes(x=LoanOriginalAmount, y=CreditScoreRangeLower,
           size=ProsperRating..numeric.,
           color=factor(IncomeRange))) + 
    
    geom_jitter() + 
    scale_color_brewer() +
    labs(title = 'Credit Score vs Loan Amount in D.C.',
         x = 'Loan Amount',
         y = 'Credit Score (Lower of Range)',
         color = 'Income Range',
         size = 'Prosper Rating')

chart3 + theme_economist(base_family="Verdana") +
  scale_colour_economist() + theme(legend.position="right")

```

### Credit Score vs Loan Amount in Washington, D.C.

> This chart takes another look at the Income and Prosper Rating in D.C. but now we are looking at the relationship to the lower range of the Credit Rating. We can see from this graph that the Prosper Rating is lower for those with a lower credit score, as we would expect, but we can also see that there are those with small loan amounts and scores around 700 that also have a low rating. The reasoning behind choosing this plot was to enable us to see if the Prosper rating was liable to correlate to certain loan amounts.

------

# Reflection

> I think that this dataset was really interesting, and the vastness of it means that I could have explored many more areas including looking at how Occupation affects other variables. I think that I learnt a lot using many resources (detailed in 'reference materials') that helped me convey my points effectively. I think that one of the aspects that I found most interesting was looking at the Income Range variable and seeing how that related back to others, and how I could again relate this to real-life examples.

>I ran into difficulties when I first looked at the data, I think this is becuase it is real-life data using loans taken out all over the US (where there is a big divide from state to state) that I could not immediately see any correlations in my first sample plot. This caused me to try and plot a lot of difference variables (not all plots are included) to try and find some that I thought were relevant.

> I think I found success once I focused on a few key variables and I really tried analysing them together, I also think that looking at certain states enabled me to see clearer patterns that were not clouded by the sheer amount of data.
